---
title: "A Tale of Two Combines"
author: "Robert A. Stevens"
date: "08/18/2017"
output: html_document
---

by Robert A. Stevens, February 24, 2014

Adpated from “Modeling and Analysis of Row Crop Harvesting Patterns by Combines” by A.C. Hansen, Q. Zhang, and T.A. Wilcox, Transactions of the ASABE, Vol. 50(1):5-12 (2007)

# Background Info:  Domain Knowledge

Table 1. Summary of equipment, field details, and electronic flags.

Figure 1. Harvesting patterns for single combine in a row crop. 

# Data

## Combine 1

1. Corn-harvesting.xls (2 tabs)

2. Corn-harvesting-CAN.csv (CAN tab)

3. Combine1.pptx

## Combine 2

1. Corn-Harvest-2.xls

2. Corn-Harvest-2-CAN.csv

3. Combine2a.pptx (as delivered – records out of time order)

4. Combine2b.pptx (fixed – records in time order)

# Variables Captured

1. LONGITUDE

2. LATITUDE

3. TIME

4. STATUS (head position:  1 = down, 0 = up)

5. DRY_YIELD (bushels/acre)

6. DAYOFMONTH

7. HOUR

8. MINUTE

9. SECOND

10. SPEED (ground speed in mph)

11. TORQUE (% of torque at rated)

12. ENG_SPD (rpm)

13. CURVE (unloading auger:  1 = off, 2 = engaged)

14. FUEL (% of maximum?)

15. FUEL_KG.HR (fuel rate in kg/hour)

16. GRD_SPD (ground speed in km/hour)

17. MPH (ground speed in mph)

Load the necessary R packages.

```{r}
library(ggplot2)    # Used for several of the plots
library(maps)       # Required for correct Lon/Lat projections for ggplot2 coord_map()
library(plyr)       # Used for sorting dataframe to make sure in the correct order
library(ggmap)      # Used for Google Earth style plots
library(rpart)      # Recursive Partitioning data mining routine
library(rpart.plot) # Improves the rpart plots
```

Load, prepare, and merge the two data sets.

```{r}
setwd("~/GitHub/Presentations") # user specific - need to edit to run
filename <- "Corn-harvesting-CAN.csv"
Combine1 <- read.csv(filename, header = TRUE, as.is = TRUE)
filename <- "Corn-Harvest-2-CAN.csv"
Combine2 <- read.csv(filename, header = TRUE, as.is = TRUE)
Combine2 <- arrange(Combine2, DAYOFMONTH, HOUR, MINUTE, SECOND) # put in the right time order
Combine1 <- Combine1[ , -c(3:4)] # remove Easting and Northing
dim(Combine1)
Combine2 <- Combine2[ , -c(11)] # remove TIMELAPSE
dim(Combine2)
Combine1$Combine <- 1
Combine2$Combine <- 2
Combine12 <- rbind(Combine1, Combine2)
Combine12$Combine <- as.factor(Combine12$Combine)
dim(Combine12)

center <- function(x) {
  (min(x, na.rm = TRUE) + max(x, na.rm = TRUE))/2
}

mapLon <- center(Combine12$LONGITUDE)
mapLat <- center(Combine12$LATITUDE)
```

The location is:

```{r}
revgeocode(c(mapLon, mapLat))
```

Use ggmap with zoom values 10 to 17 to get an idea of where this is:

```{r ggplot, fig.width = 10, fig.height = 7.5}
plotTitle <- "Combine 1 and Combine 2 Paths using ggmap with zoom = 10"
map <- get_googlemap(center = c(lon = mapLon, lat = mapLat), zoom = 10, maptype = 'hybrid')
ggmap(map) + 
  geom_path(data = Combine12, aes(x = LONGITUDE, y = LATITUDE, color = Combine), size = I(1)) +
  labs(title = paste(plotTitle), x = "Longitude", y = "Latitude")

plotTitle <- "Combine 1 and Combine 2 Paths using ggmap with zoom = 11"
map <- get_googlemap(center = c(lon = mapLon, lat = mapLat), zoom = 11, maptype = 'hybrid')
ggmap(map) + 
  geom_path(data = Combine12, aes(x = LONGITUDE, y = LATITUDE, color = Combine), size = I(1)) +
  labs(title = paste(plotTitle), x = "Longitude", y = "Latitude")
makeFootnote(footnote)

plotTitle <- "Combine 1 and Combine 2 Paths using ggmap with zoom = 12"
map <- get_googlemap(center = c(lon = mapLon, lat = mapLat), zoom = 12, maptype = 'hybrid')
ggmap(map) + 
  geom_path(data = Combine12, aes(x = LONGITUDE, y = LATITUDE, color = Combine), size = I(1)) +
  labs(title = paste(plotTitle), x = "Longitude", y = "Latitude")

plotTitle <- "Combine 1 and Combine 2 Paths using ggmap with zoom = 13"
map <- get_googlemap(center = c(lon = mapLon, lat = mapLat), zoom = 13, maptype = 'hybrid')
ggmap(map) + 
  geom_path(data = Combine12, aes(x = LONGITUDE, y = LATITUDE, color = Combine), size = I(1)) +
  labs(title = paste(plotTitle), x = "Longitude", y = "Latitude")

plotTitle <- "Combine 1 and Combine 2 Paths using ggmap with zoom = 14"
map <- get_googlemap(center = c(lon = mapLon, lat = mapLat), zoom = 14, maptype = 'hybrid')
ggmap(map) + 
  geom_path(data = Combine12, aes(x = LONGITUDE, y = LATITUDE, color = Combine), size = I(1)) +
  labs(title = paste(plotTitle), x = "Longitude", y = "Latitude")

plotTitle <- "Combine 1 and Combine 2 Paths using ggmap with zoom = 15"
map <- get_googlemap(center = c(lon = mapLon, lat = mapLat), zoom = 15, maptype = 'hybrid')
ggmap(map) + 
  geom_path(data = Combine12, aes(x = LONGITUDE, y = LATITUDE, color = Combine), size = I(1)) +
  labs(title = paste(plotTitle), x = "Longitude", y = "Latitude")

plotTitle <- "Combine 1 and Combine 2 Paths using ggmap with zoom = 16"
map <- get_googlemap(center = c(lon = mapLon, lat = mapLat), zoom = 16, maptype = 'hybrid')
ggmap(map) + 
  geom_path(data = Combine12, aes(x = LONGITUDE, y = LATITUDE, color = Combine), size = I(1)) +
  labs(title = paste(plotTitle), x = "Longitude", y = "Latitude")

plotTitle <- "Combine 1 and Combine 2 Paths using ggmap with zoom = 17"
map <- get_googlemap(center = c(lon = mapLon, lat = mapLat), zoom = 17, maptype = 'hybrid')
ggmap(map) + 
  geom_path(data = Combine12, aes(x = LONGITUDE, y = LATITUDE, color = Combine), size = I(1)) +
  labs(title = paste(plotTitle), x = "Longitude", y = "Latitude")
```

Explore the data using combine paths and color coding for other variables:

```{r fig.width = 10, fig.height = 7.5}
Combine12$Combine <- as.factor(Combine12$Combine)
plotTitle <- "Combine 1 and Combine 2 Paths"
qplot(LONGITUDE, LATITUDE, data = Combine12, geom = "point", color = Combine, size=I(2)) +
  scale_x_continuous('Longitude') + 
  scale_y_continuous('Latitude') + 
  coord_map() +
  labs(title = paste(plotTitle), x = "Longitude", y = "Latitude")

plotTitle <- "Harvest Pattern for Combines 1 and 2 with CURVE (unloading auger:  1 = off, 2 = engaged)"
Combine12$CURVE <- as.factor(Combine12$CURVE)
qplot(LONGITUDE, LATITUDE, data = Combine12, geom = "point", color = CURVE, size = I(2)) +
  scale_x_continuous('Longitude') + 
  scale_y_continuous('Latitude') + 
  coord_map() +
  labs(title = paste(plotTitle)) +
  facet_grid(. ~ Combine)

plotTitle <- "Harvest Pattern for Combines 1 and 2 with STATUS (head position:  1 = down, 0 = up)"
Combine12$STATUS <- as.factor(Combine12$STATUS)
qplot(LONGITUDE, LATITUDE, data = Combine12, geom = "point", color = STATUS, size = I(2)) +
  scale_x_continuous('Longitude') + 
  scale_y_continuous('Latitude') + 
  coord_map() +
  labs(title = paste(plotTitle)) +
  facet_grid(. ~ Combine)

plotTitle <- "Harvest Pattern for Combines 1 and 2 with Dry Yield (bushels/acre)"
qplot(LONGITUDE, LATITUDE, data=Combine12, geom="point", color=DRY_YIELD, size=I(2)) +
  scale_x_continuous('Longitude') + 
  scale_y_continuous('Latitude') + 
  coord_map() +
  labs(title = paste(plotTitle)) +
  facet_grid(. ~ Combine) +
  scale_colour_gradientn(colours = rainbow(7))

plotTitle <- "Dry Yield Histograms (Combine 1 = Red, Combine 2 = Blue)"
ggplot(Combine12, aes(x = DRY_YIELD)) + 
  geom_histogram(data=subset(Combine12, Combine == 1), fill = "red", alpha = 0.2) +
  geom_histogram(data=subset(Combine12, Combine == 2), fill = "blue", alpha = 0.2) +
  labs(title = paste(plotTitle), x = "Dry Yield (bushels/acre)") + 
  theme_bw()

plotTitle <- "Harvest Pattern for Combines 1 and 2 with Engine Speed (rpm)"
qplot(LONGITUDE, LATITUDE, data = Combine12, geom = "point", color = ENG_SPD, size = I(2)) +
  scale_x_continuous('Longitude') + 
  scale_y_continuous('Latitude') + 
  coord_map() +
  labs(title = paste(plotTitle)) +
  facet_grid(. ~ Combine) +
  scale_colour_gradientn(colours = rainbow(7))

plotTitle <- "Engine Speed Histograms (Combine 1 = Red, Combine 2 = Blue)"
ggplot(Combine12, aes(x=ENG_SPD)) + 
  geom_histogram(data=subset(Combine12, Combine == 1), fill = "red", alpha = 0.2) +
  geom_histogram(data=subset(Combine12, Combine == 2), fill = "blue", alpha = 0.2) +
  labs(title = paste(plotTitle), x = "Engine Speed (rpm)") + 
  theme_bw() +
  xlim(2100, 2500)

plotTitle <- "Harvest Pattern for Combines 1 and 2 with TORQUE (% of torque at rated)"
qplot(LONGITUDE, LATITUDE, data = Combine12, geom = "point", color = TORQUE, size = I(2)) +
  scale_x_continuous('Longitude') + 
  scale_y_continuous('Latitude') + 
  coord_map() +
  labs(title = paste(plotTitle)) +
  facet_grid(. ~ Combine) +
  scale_colour_gradientn(colours = rainbow(7))

plotTitle <- "TORQUE Histograms (Combine 1 = Red, Combine 2 = Blue)"
ggplot(Combine12, aes(x = TORQUE)) + 
  geom_histogram(data=subset(Combine12, Combine == 1), fill = "red", alpha = 0.2) +
  geom_histogram(data=subset(Combine12, Combine == 2), fill = "blue", alpha = 0.2) +
  labs(title = paste(plotTitle), x = "TORQUE (% of torque at rated)") + 
  theme_bw()

plotTitle <- "Harvest Pattern for Combines 1 and 2 with Ground Speed (mph)"
qplot(LONGITUDE, LATITUDE, data = Combine12, geom = "point", color = MPH, size = I(2)) +
  scale_x_continuous('Longitude') + 
  scale_y_continuous('Latitude') + 
  coord_map() +
  labs(title = paste(plotTitle)) +
  facet_grid(. ~ Combine) +
  scale_colour_gradientn(colours = rainbow(7))

plotTitle <- "Ground Speed Histograms (Combine 1 = Red, Combine 2 = Blue)"
ggplot(Combine12, aes(x=MPH)) + 
  geom_histogram(data=subset(Combine12, Combine == 1), fill = "red", alpha = 0.2) +
  geom_histogram(data=subset(Combine12, Combine == 2), fill = "blue", alpha = 0.2) +
  labs(title = paste(plotTitle), x = "Ground Speed (mph)") + 
  theme_bw()

plotTitle <- "Harvest Pattern for Combines 1 and 2 with Fuel Rate (kg/hour)"
qplot(LONGITUDE, LATITUDE, data = Combine12, geom = "point", color = FUEL_KG.HR, size = I(2)) +
  scale_x_continuous('Longitude') + 
  scale_y_continuous('Latitude') + 
  coord_map() +
  labs(title = paste(plotTitle)) +
  facet_grid(. ~ Combine) +
  scale_colour_gradientn(colours = rainbow(7))

plotTitle <- "Fuel Rate Histograms (Combine 1 = Red, Combine 2 = Blue)"
ggplot(Combine12, aes(x = FUEL_KG.HR)) + 
  geom_histogram(data = subset(Combine12, Combine == 1), fill = "red", alpha = 0.2) +
  geom_histogram(data = subset(Combine12, Combine == 2), fill = "blue", alpha = 0.2) +
  labs(title = paste(plotTitle), x = "Fuel Rate (kg/hour)") + 
  theme_bw()
```

Now try some data mining with Recursive Partitioning (rpart):

```{r fig.width = 10, fig.height = 7.5}
FuelRate.rp <- rpart(FUEL_KG.HR ~ STATUS + DRY_YIELD + TORQUE + ENG_SPD + CURVE + MPH + Combine, 
                     data = Combine12, cp = 1e-3)
plotcp(FuelRate.rp)
```

The last node just crosses the threhold, so no need to prune the tree.

```{r}
FuelRate.rp
```

Note:  the only variables that appear are TORQUE, ENG_SPD and CURVE (unloading auger:  1 = off, 2 = engaged).

```{r fig.width = 10, fig.height = 7.5}
plotTitle <- "rpart Results for Fuel Rate (kg/hour)"
plot(FuelRate.rp, main = plotTitle)
text(FuelRate.rp, digits = 4, cex = 0.65)
```

That's hard to read, so try "prp" (pretty rpart):

```{r fig.width = 10, fig.height = 7.5}
plotTitle <- "rpart Results for Fuel Rate (kg/hour) using prp"
prp(FuelRate.rp, main = plotTitle)
```

That's better, but still difficult to explain, so try enhancing prp:

```{r fig.width = 10, fig.height = 7.5}
plotTitle <- "rpart Results for Fuel Rate (kg/hour) using prp Enhanced"
prp(FuelRate.rp, type = 2, nn = TRUE, fallen.leaves = TRUE, faclen = 0, varlen = 0, shadow.col = "grey", branch.lty = 3, 
    main = plotTitle)
```

Now plot all 4 variables on one plot:  ENG_SPD, TORQUE, CURVE and Fuel Rate.  

Played around with ggplot2's "alpha" and "size" to create a pseudo contour plot.

```{r fig.width = 10, fig.height = 7.5}
plotTitle <- "ENG_SPD versus TORQUE by CURVE (unloading auger:  1 = off, 2 = engaged)"
qplot(ENG_SPD, TORQUE, data = Combine12, color = FUEL_KG.HR, alpha = I(1/4), size = I(4)) + 
  facet_grid(. ~ CURVE) +
  labs(title = paste(plotTitle)) +
  scale_colour_gradientn(colours = rainbow(7)) +
  theme_bw()
makeFootnote(footnote)
```

Even though Combine didn't show up in the data mining results, put in in anyway to see what it looks like.

```{r fig.width = 10, fig.height = 7.5}
plotTitle <- "ENG_SPD versus TORQUE by CURVE (columns) and Combine (rows)"
qplot(ENG_SPD, TORQUE, data = Combine12, color = FUEL_KG.HR, alpha = I(1/4), size = I(4)) + 
  facet_grid(Combine ~ CURVE) +
  labs(title = paste(plotTitle)) +
  scale_colour_gradientn(colours = rainbow(7)) +
  theme_bw()
```

It seems the two combines performed differently because they had different levels of Engine Speed and Torque - why?

# Still work to do

- Calculate productivity metrics (e.g. tons/hour):  Need to “normalize” the data

- Add Elevation/Altitude and other environmental variables

- Predictive Modeling – interactions?  nonlinear?

- Animations:  R, Tableau, other…

## All along the way, review results with engineers:

- Why are Combines 1 and 2 different?  Machine or Operator?

- Or maybe there were data collection errors?

```{r}
citation('ggmap')
```
