# University of Illinois at Urbana-Champaign Research Park Big Data Summit

by Robert A. Stevens, John Deere, December 6, 2013

The information, views, and opinions contained in this presentation are those of the author and do not necessarily reflect the views and opinions of John Deere.

# What is Analytics?  Turning Data into Decisions

## Analytics is the application of information technology, statistics, and decision sciences to turn data into actionable information that solves a business problem.

```{r fig.width = 10, fig.height = 7.6, echo = FALSE, warning = FALSE}
library(SixSigma, quietly = TRUE)
inputs <-c ("Data")
outputs <- c("Decisions")
steps <- c("IT", "Statistics", "Decision Sciences")
io <- list()
io[[1]] <- list("Servers", "Databases", "Software", "etc.")
io[[2]] <- list("Data Mining", "Machine Learning", "Modeling", "etc.")
io[[3]] <- list("Framing", "Modeling", "Optimizing", "etc.")
param <- list()
param[[1]] <- list(c("Technology", "C"), c("+ People", "C"), c("= Capability", "Cr"))
param[[2]] <- list(c("Technology", "C"), c("+ People", "C"), c("= Capability", "Cr"))
param[[3]] <- list(c("Technology", "C"), c("+ People", "C"), c("= Capability", "Cr"))
feat <- list()
feat[[1]] <- list("Cost", "Timing", "Quality")
feat[[2]] <- list("Cost", "Timing", "Quality")
feat[[3]] <- list("Cost", "Timing", "Quality")
ss.pMap(steps, inputs, outputs, io, param, feat, sub = "Analytics")
```

## “It’s not the figures themselves,” she said finally, “it’s what you do with them that matter.”

From *Advanced Theory of Statistics Vol.2* by Kendall & Stuart and *Repairable Systems Reliability* by Ascher & Feingold

**You must TAKE ACTION to realize any value from all the time spent on IT, Statistics and Decision Sciences.**

# The Road to Earlier Discovery and Shorter Decision Cycles

```{r ggplot, fig.width = 10, fig.height = 7.5, echo=FALSE}
library(ggplot2, quietly = TRUE)
Phase <- c(rep("Today", 3), rep("Transition", 3), rep("Vision", 3))
Activity <- rep(c("IT", "Statistics", "Decisions"), 3)
Activity <- ordered(Activity, c("IT", "Statistics", "Decisions"))
Color <- rep(c("red", "yellow", "green"), 3)
Time <- c(70, 20, 10, 33.3, 33.3, 33.3, 10, 20, 70)
plotDF <- data.frame(Phase, Activity, Color, Time)
qplot(Activity, Time, data = plotDF, geom = "bar", stat="identity", fill = Activity) + 
  facet_wrap(~ Phase) +
  labs(x = "Analytics Activity", y = "Time (%)") +
  scale_fill_manual(values=Color) +
  theme(legend.position = "none")
```

## Today - Primarily IT focused

* Begin to enhance IT with Statistics and Decision Sciences

## Transition - More Statistics and Decision Sciences

* IT technology improves and costs drop

* Data Mining to discover relationships

* Decisions start to drive IT and Statistics

## Vision - Primarily Decision focused

* Low cost, automated IT

* Machine Learning to automate statistics

* Humans spend most of their time on decisions

**How long will this journey take?  I've been told it took FEA/CFD 20 years to get to where they are today.  Let's hope we can accelerate our rate of progress.**

# Big Data in R&D

## Primarily machine data:  GPS and CAN (Controller Area Network)

* Volume:  immeasurable, but we don’t keep most because we cannot afford to; the more we can afford to keep, the more we will keep

* Velocity:  30 minutes (JDLink); 10 millisecond to 1 second (MGDL); 100,000 per second (eDAQ)

* Variety:  different sampling rates; same thing with different names; different things with same name

* Value:  TBD – that’s my job

**Unfortunately, in R&D all Statistical and Decision Science work is done on PCs**

Reference:  http://en.wikipedia.org/wiki/CAN_bus

# Goal:  Move Statistical and Decision Science work off of PCs and onto a server

## Big Data Studies/Solutions

* IBM Netezza with Revolution R

* IBM BigInsights (Hadoop) with R (requires JAQL)

* SAP HANA with R

* Cloudera/Hortonworks Hadoop with R

* Linux cluster with Hadoop, R, RHIPE, BigR, etc.

* Savvis Cloud???

## And consequently, I get invited to IT meetings that I never did before because nobody knows what R is

**Our original strategy was to select ONE (the *Highlander* approach), but now “we” realize there must be a Big Data Ecosystem (a.k.a. a System of Systems)**

# My work/life philosophy

## Old:  A good day at work is being able to do statistics

* Very rare

## New:  A good day at work is being able to use R

* Becoming more common – Thank you, Big Data!

# R and Big Data Personal Observations from a Business Perspective

## Principal-Agent Problem Example:  Empire-building

* Old:  adding staff to increase your org chart

* New:  adding expensive, proprietary hardware and software (and staff to run them) to increase your budget and span of control

## Moral Hazard Problem:

* Asking and accepting advice from the same people whose career and career path depend on the answer they provide

## Rhetorical Question:

* What would a shareholder or an analyst think if they found out you spent 10X, 20X, or even 100X for hardware or software instead of a Linux cluster with open source software that offered the same or even more benefits?

References:

http://en.wikipedia.org/wiki/Principal%E2%80%93agent_problem

http://en.wikipedia.org/wiki/Empire-building

# Contact Information

* Business e-mail stevensroberta@johndeere.com

* Personal e-mail	robertandrewstevens@gmail.com

* LinkedIn http://www.linkedin.com/pub/robert-andrew-stevens-cfa/6a/a04/315

* Twitter 	https://twitter.com/RobertAndrewSt3

* GitHub https://github.com/robertandrewstevens
